---
title: "changeOD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{changeOD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = FALSE
)
```

```{r setup, eval=TRUE}
library(collapse)
library(ggplot2)
```

# Introduction

This document provides an introduction to 2011 and 2021 OD data and how they can be compared to explore changes in travel patterns in relation to particular corridors affected by new interventions.

The reproducible source code underlying this document can be found at [github.com/robinlovelace/changeOD](https://github.com/robinlovelace/changeOD/blob/main/vignettes/changeOD.Rmd).

# Datasets

## 2011 Origin-Destination data

2011 data was taken from the open access data provided by the WICID service and downloaded with the function `get_od` from the `pct` package. The data is available at [WICID](https://wicid.ukdataservice.ac.uk/flowdata/cider/wicid/downloads.php) and is based on the 2011 Census. The data is available at the local authority level and includes the following variables (after removing unwanted variables):

- `geo_code1`: Origin code
- `geo_code2`: Destination code
- `from_home`: Working from home
- `light_rail`: Underground, metro, light rail, tram
- `train`: Train
- `bus`: Bus
- `taxi`: Taxi
- `motorbike`: Motorcycle, scooter or moped
- `car_driver`: Driving a car or van
- `car_passenger`: Passenger in a car or van
- `bicycle`: Bicycle
- `foot`: On foot
- `other`: Other method of travel to work

```{r, eval=FALSE}
od = pct::get_od()
names(od)
#  [1] "geo_code1"     "geo_code2"     "all"           "from_home"
#  [5] "light_rail"    "train"         "bus"           "taxi"
#  [9] "motorbike"     "car_driver"    "car_passenger" "bicycle"
# [13] "foot"          "other"         "geo_name1"     "geo_name2"
# [17] "la_1"          "la_2"
# Remove unwanted variables with fselect:
od_2011_minimal = od |>
  fselect(geo_code1, geo_code2, all)
arrow::write_parquet(od_2011_minimal, "od_2011_minimal.parquet")
od = od |>
  fselect(-la_1, -la_2, -geo_name1, -geo_name2, -all, -from_home)
# Save as arrow file:
arrow::write_parquet(od, "od_2011.parquet")
# Upload to GitHub with piggyback:
piggyback::pb_upload("od_2011.parquet")
piggyback::pb_upload("od_2011_minimal.parquet")
```

```{r}
od_2011 = arrow::read_parquet("od_2011.parquet")
```

```{r, eval=FALSE}

```

## 2021 Origin-Destination data

See the data at [WICID](https://wicid.ukdataservice.ac.uk/flowdata/cider/wicid/downloads.php) (requires login for data with mode).

The open access OD data can be obtained as follows from this link: [WICID](https://wicid.ukdataservice.ac.uk/cider/wicid/downloads.php#).

```{r, eval=FALSE}
od_2021_open_raw = readr::read_csv("ODWP12EW_MSOA_v1.zip")
names(od_2021_open_raw)
# [1] "Middle layer Super Output Areas code"        
# [2] "Middle layer Super Output Areas label"       
# [3] "MSOA of workplace code"                      
# [4] "MSOA of workplace label"                     
# [5] "Hours worked (5 categories) code"            
# [6] "Hours worked (5 categories) label"           
# [7] "Place of work indicator (4 categories) code" 
# [8] "Place of work indicator (4 categories) label"
# [9] "Count"  
# Table of place of work indicator:
table(
  od_2021_open_raw$`Place of work indicator (4 categories) label`
)
# Pivot wider and create new 'all' column:
od_2021_open = od_2021_open_raw |>
  subset(
    `Place of work indicator (4 categories) label` == "Working in the UK but not working at or from home" |
    `Place of work indicator (4 categories) label` == "Mainly working at or from home" |
    `Place of work indicator (4 categories) label` == "Other (including offshore installation, working outside the UK)"
  ) |>
  fgroup_by(
    `Middle layer Super Output Areas code`,
    `MSOA of workplace code`
  ) |>
  fsummarise(all = sum(Count))
names(od_2021_open) = c(
  "geo_code1",
  "geo_code2",
  "all"
)
sum(od_2021_open$all)
# 15 million trips
# Save as arrow file:
arrow::write_parquet(od_2021_open, "od_2021_open.parquet")
# Upload the file:
piggyback::pb_upload("od_2021_open.parquet")
```

We undertook the following steps to process the data:

1. Download the data from the WICID service.
2. Read the data into R.
3. Check the structure of the data.
4. Check the cross-tabulation of mode
5. Aggregate the data by mode of travel and remove unwanted places of work.
6. Pivot the data wider to match the 2011 dataset.
7. Change the names to match the 2011 dataset.
8. Compare the 2021 and 2011 datasets.

```{r, eval=FALSE}
list.files(pattern = "zip")
od_2022_raw = readr::read_csv("ODWP15EW_MSOA_v1.zip")
names(od_2022_raw)
names(od_2022_raw)
#  [1] "Middle layer Super Output Areas code"
#  [2] "Middle layer Super Output Areas label"
#  [3] "MSOA of workplace code"
#  [4] "MSOA of workplace label"
#  [5] "Method used to travel to workplace (12 categories) code"
#  [6] "Method used to travel to workplace (12 categories) label"
#  [7] "Sex (2 categories) code"
#  [8] "Sex (2 categories) label"
#  [9] "Age (E) (8 categories) code"
# [10] "Age (E) (8 categories) label"
# [11] "Place of work indicator (4 categories) code"
# [12] "Place of work indicator (4 categories) label"
# [13] "Count"
# Check cross-tabulation of mode of travel with Sex:
table(
  od_2022_raw$`Method used to travel to workplace (12 categories) label`,
  od_2022_raw$`Sex (2 categories) label`
)
# And Age:
table(
  od_2022_raw$`Method used to travel to workplace (12 categories) label`,
  od_2022_raw$`Age (E) (8 categories) label`
)
# Age and sex:
table(
  od_2022_raw$`Age (E) (8 categories) label`,
  od_2022_raw$`Sex (2 categories) label`
)
# Just place of work:
table(
  od_2022_raw$`Place of work indicator (4 categories) label`
)


#                                                  Does not apply 
#                                                           87168 
#                  Mainly working at or from home, No fixed place 
#                                                          510823 
# Other (including offshore installation, working outside the UK) 
#                                                           66219 
#               Working in the UK but not working at or from home 
#                                                         7097394 
# And all 4:
table(
  od_2022_raw$`Method used to travel to workplace (12 categories) label`,
  od_2022_raw$`Sex (2 categories) label`,
  od_2022_raw$`Age (E) (8 categories) label`,
  od_2022_raw$`Place of work indicator (4 categories) label`
)
sum(od_2022_raw$Count)
# Aggregating by mode of travel after removing unwanted places of work:
od_2022_aggregated = od_2022_raw |>
  # subset(
  #   `Place of work indicator (4 categories) label` == "Working in the UK but not working at or from home" |
  #   `Place of work indicator (4 categories) label` == "Mainly working at or from home, No fixed place"
  #   ) |>
  fgroup_by(
    `Middle layer Super Output Areas code`,
    `MSOA of workplace code`,
    `Method used to travel to workplace (12 categories) label`
  ) |>
  fsummarise(Count = sum(Count))
# Pivot wider keeping only MSOA code (geo_code new name) and mode of travel (mode new name):
# od_2022_by_mode = od_2022_raw |> # generates incorrect results
od_2022_by_mode = od_2022_aggregated |>
  pivot(
    id = c(
      "Middle layer Super Output Areas code",
      "MSOA of workplace code"
    ),
    values = "Count",
    names = "Method used to travel to workplace (12 categories) label",
    how = "wider"
  )
names(od_2022_by_mode)
#  [1] "Middle layer Super Output Areas code"        
#  [2] "MSOA of workplace code"                      
#  [3] "Not in employment or aged 15 years and under"
#  [4] "Bicycle"                                     
#  [5] "Bus, minibus or coach"                       
#  [6] "Driving a car or van"                        
#  [7] "On foot"                                     
#  [8] "Other method of travel to work"              
#  [9] "Passenger in a car or van"                   
# [10] "Taxi"                                        
# [11] "Train"                                       
# [12] "Underground, metro, light rail, tram"        
# [13] "Motorcycle, scooter or moped"                
# [14] "Work mainly at or from home"            

# Change names to match 2011 dataset:
names(od_2022_by_mode) = c(
  "geo_code1",
  "geo_code2",
  "not_in_employment",
  "bicycle",
  "bus",
  "car_driver",
  "foot",
  "other",
  "car_passenger",
  "taxi",
  "train",
  "light_rail",
  "motorbike",
  "from_home"
)
ncol(od_2022_by_mode)
ncol(od_2011)
waldo::compare(names(od_2022_by_mode), names(od_2011))
od_2021 = od_2022_by_mode[names(od_2011)]
# Set all NAs to 0:
od_2021[is.na(od_2021)] = 0
# Compare working from home numbers:
sum(od_2011$from_home)
sum(od_2021$from_home) # ~4 x more people working from home
# Driving:
sum(od_2011$car_driver)
sum(od_2021$car_driver)
# Walking:
sum(od_2011$foot)
sum(od_2021$foot)
# Save as arrow file:
arrow::write_parquet(od_2021, "od_2021.parquet")
```

```{r}
od_2021 = arrow::read_parquet("od_2021.parquet")
```

```{r}
# Check the summaries are similar:
cor(
  colMeans(od_2011[-(1:2)], na.rm = TRUE),
  colMeans(od_2021[-(1:2)], na.rm = TRUE)
)
# [1] 0.8809119
# Save as combined dataset:
od_combined = rowbind(
  od_2011 |> fmutate(year = 2011),
  od_2021 |> fmutate(year = 2021)
)
# Save as arrow file:
arrow::write_parquet(od_combined, "od_combined.parquet")
```

# OD basics

Basic info on the 2011 and 2021 datasets are presented in the table below.

```{r}
od_combined = arrow::read_parquet("od_combined.parquet")
od_summary = data.frame(
  year = c(2011, 2021),
  n_row = c(nrow(od_2011), nrow(od_2021)),
  sum = c(sum(od_2011[-(1:2)]), sum(od_2021[-(1:2)]))
)
# od_summary |>
#   knitr::kable()
```

| year|   n_row|      sum|
|----:|-------:|--------:|
| 2011| 2402201| 26681568|
| 2021| 1851505| 27774281|

The change from 2011 to 2021 is shown in the figure below

```{r}
od_mode_sums = od_combined |>
  fgroup_by(year) |>
  fsummarise(across(-c(year, geo_code1, geo_code2), sum)) |>
  pivot("year") |>
  # Mutate names to be in title case, e.g. car_driver -> Car driver
  fmutate(variable = tools::toTitleCase((as.character(variable)))) |>
  # Replace _ with space
  fmutate(variable = gsub("_", " ", variable))
# Manually set order of modes:
# new_levels = c()
names(od_mode_sums) = c("year", "Mode", "Number of trips")
# Save as arrow file:
arrow::write_parquet(od_mode_sums, "od_mode_sums.parquet")
piggyback::pb_upload("od_mode_sums.parquet")
# get url:
```

```{r, eval=TRUE}
u = "https://github.com/Robinlovelace/changeOD/releases/download/latest/od_mode_sums.parquet"
od_mode_sums = arrow::read_parquet(u)
```

```{r}
```

```{r}
```

```{r}
```

```{r, eval=TRUE}
# Plot in bar chart:
od_mode_sums |>
  ggplot(
    aes(x = Mode, y = `Number of trips`, fill = factor(year))
  ) +
  geom_col(position = "dodge") +
  theme_minimal() +
  # Human readable y axis labels:
  scale_y_continuous(
    labels = scales::comma
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    x = "Mode of travel",
    y = "Number of trips",
    fill = "Year"
  )
```
